\documentclass[12pt]{article}

% House style preamble
\input{.house-style/preamble.tex}

% Update PDF metadata
\hypersetup{
  pdftitle={English Interjections as a Homeostatic Property Cluster}
}

\title{English Interjections as a Homeostatic Property Cluster}
\author{Brett Reynolds \orcidlink{0000-0003-0073-7195}%
\thanks{Contact: \href{mailto:brett.reynolds@humber.ca}{brett.reynolds@humber.ca}}\\
Humber Polytechnic \& University of Toronto}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
English interjections have been denied category status
\citep{jespersen1924}, dismissed as uninteresting
\citep{HuddlestonPullum2005}, or acknowledged as \enquote{marginal and
anomalous} \citep[67]{quirk1985}. This paper argues that interjections
are a thin-but-genuine homeostatic property cluster (HPC): a set of
co-occurring properties (prosodic isolation, non-inflection, syntactic
non-integration, non-referentiality, emotive force) maintained by
identifiable stabilisers. Multiple diachronic pathways~--
grammaticalization via semantic bleaching, pragmaticalization via
syntactic deactivation, and nonce coinage via onomatopoeia~-- converge
on the same synchronic cluster, extending \textcite{gehweiler2008}'s
analysis of \mention{gee!} to a general theoretical claim. A
perturbation test suggests that removing a stabiliser produces
predictable, asymmetric category shifts. The paper's distinctive
contribution is that interjection projectibility is primarily pragmatic
rather than truth-conditional: the most frequent interjections are
interactional tools (continuers, repair initiators, change-of-state
tokens), and categorizing a novel form as an interjection licenses
inferences about prosodic behaviour, interactional distribution, and
stance expression. A field-relative decomposition
shows that syntacticians, semanticists, and interactional linguists
each track distinct HPCs with overlapping but non-identical extensions.
The category that embarrassed classical theory is a natural fit for
HPC~-- and its pragmatic projectibility extends the framework to a new
domain of inference.
\end{abstract}

\section{Introduction}\label{sec:intro}

In 1586, William Bullokar wrote the earliest grammar of English. It
included interjections. Bullokar defined them as words that betoken
\enquote{a sudden passion of the mind}
\citep[373]{bullokar1586}~-- a definition that foregrounds meaning and
use rather than distributional properties.

Four centuries later, the category's status has declined.
\textcite[90]{jespersen1924} treated interjections not as a lexical
category but as a manner in which words of other categories may be used.
\textcite[67]{quirk1985} included the category but called it
\enquote{marginal and anomalous.}
\textcite[22]{HuddlestonPullum2002} listed interjection as one of nine
lexical categories in their comprehensive grammar (\mention{ah},
\mention{damn}, \mention{gosh}, \mention{hey}, \mention{oh},
\mention{ooh}, \mention{ouch}, \mention{whoa}, \mention{wow}), but
\textcite[16]{HuddlestonPullum2005} dropped it from the textbook
because \enquote{there really isn't anything interesting for a grammar
to say} about interjections. A category real enough for the definitive
grammar but thin enough for the textbook to cut.

The empirical picture, though, hasn't simplified. English interjections
span phonology, morphology, syntax, semantics, and pragmatics. The
category resists clean definition~-- no single property is necessary
and sufficient~-- and its boundaries with nouns, verbs, adverbs,
fillers, and routine formulae are graded, not sharp.
Cross-linguistically, \enquote{interjection} is at best a comparative
concept \citep{haspelmath2010}. This paper asks not \enquote{are
interjections a category?} but \enquote{what does categorizing something
as an interjection let you predict?} A category that supports no
inferences is a filing label. A category that supports reliable
inferences across novel instances is a genuine kind.

A methodological note: HPC membership, as used here, attaches to
constructions (uses), not to lexemes. \mention{Damn} as a standalone
interjection, \mention{damn} as an attributive modifier in \mention{the
damn car}, and \mention{damn} as a verb in \mention{I damn you} are
different constructions with different cluster profiles~-- they may
belong to different HPCs or to the same HPC in different degrees.
Lexemes serve as convenient labels; the unit of analysis is the
construction.

Section~\ref{sec:cluster} documents the cluster.
Section~\ref{sec:hpc} introduces the theoretical framework~-- homeostatic
property cluster (HPC) theory \citep{boyd1999}~-- and explains why it
handles the interjection data better than classical or prototype-based
alternatives. The rest of the paper identifies the
mechanisms that sustain the cluster, extending
\textcite{gehweiler2008}'s single-case analysis (published in this
journal) to a general theoretical claim, and argues that interjections
exhibit a distinctive form of projectibility~-- primarily pragmatic
rather than truth-conditional.

\section{The property cluster}\label{sec:cluster}

English interjections exhibit a cluster of co-occurring properties across
phonology, morphology, syntax, semantics, and pragmatics. This section
surveys those properties, drawing primarily on the empirical base documented
in \textcite{ameka1992}, \textcite{quirk1985}, and
\textcite{HuddlestonPullum2002}. For each property, two questions are
asked: does it co-occur with the others, and does observing it in a novel
form license inference to the rest of the cluster?

\subsection{Phonology}

Interjections are prosodically isolated. They typically occupy their own
intonation unit, separated from surrounding discourse by pauses
\citep[108]{ameka1992}. In writing, this isolation surfaces as punctuation~--
commas, dashes, exclamation marks~-- setting the interjection off from the
clause it accompanies \citep[1350]{HuddlestonPullum2002}.

Interjections may also contain phonemes not otherwise found in the
speaker's dialect. The interjection \mention{uh-oh} ([\ipa{əʔo}]) is one of
the few cases where a glottal stop appears in dialects that otherwise lack
it. Other atypical sounds include the alveolar clicks in
\mention{tut-tut} ([\ipa{ǀǀ}]), the voiceless bilabial fricative in
\mention{whew} ([\ipa{ɸɪu}]), and~-- for dialects that no longer use it~--
the voiceless velar fricative in \mention{ugh} ([\ipa{əx}])
\citep[853]{quirk1985}. Spelling pronunciations often arise for these forms
(e.g., [\ipa{tət tət}] for \mention{tut-tut}, [\ipa{əɡ}] for
\mention{ugh}), suggesting that the atypical phonemes are marked even for
native speakers.

\textit{Inferential contribution.} Prosodic isolation is a strong cue: if
a novel form occupies its own intonation unit and resists integration into
the surrounding prosodic contour, a listener can predict syntactic
non-integration and non-referentiality. The atypical phonemes reinforce this
inference. But prosodic isolation isn't unique to interjections~-- vocatives,
appositives, and other supplements share it \citep[cf.][]{wilkins1992}. What
distinguishes interjections is the convergence of prosodic isolation with
the properties below.

\subsection{Morphology}

English interjections characteristically don't inflect or derive. They
resist the morphological operations that characterize other open-class
categories: no plural (*\mention{ouches}), no tense (*\mention{ouched}),
no comparative (*\mention{oucher}), no nominalization
(*\mention{ouchness}) \citep[106]{ameka1992}. A caveat:
\textcite{libert2019} notes that \enquote{few, if any, papers focus on
the morphology of interjections}, so the apparent absence of
morphological productivity may partly reflect a gap in the research
rather than a firm empirical generalization.

Some interjections retain remnant morphology from earlier categorial
membership. \mention{Heavens} and \mention{bollocks} preserve the nominal
plural suffix, and \mention{goddammit} preserves a verb-object-pronoun
structure, but none of these remnants are productive. The plural in
\mention{heavens} doesn't contrast with a singular *\mention{heaven!} (as
an interjection), and speakers don't extend the pattern to new forms.

\textit{Inferential contribution.} Non-inflection predicts
non-argumenthood. Forms that don't inflect can't agree with a verb or take
a determiner, so they can't fill the syntactic slots that require these
operations. But non-inflection alone isn't diagnostic: adverbs also don't
inflect \citep[106]{ameka1992}. The distinction is syntactic:
adverbs form constituents with other words, while interjections don't
\citep[152]{meinard2015}.

\subsection{Syntax}

Interjections characteristically function as \term{supplements}~--
\enquote{parenthetical strings that are not integrated in clause
structure} \citep[1350]{HuddlestonPullum2002}. In \mention{Damn, we're going to be late!}, the
interjection isn't a complement or modifier of anything in the clause.
It's appended to the beginning as a parenthetical.

Supplement function isn't unique to interjections. Relative clauses, noun
phrases, adjective phrases, and adverb phrases can all function as
supplements \citep[1356]{HuddlestonPullum2002}:

\ea\label{ex:supplements}
  \ea \mention{Ah}, so you were there after all! \hfill (interjection)
  \ex We called in to see Sue's parents, \mention{which made us rather late}. \hfill (relative clause)
  \ex A university professor, \mention{Dr.\ Brown}, was arrested. \hfill (NP)
  \z
\z
What distinguishes interjection supplements is their combination with
the other cluster properties: non-referentiality, non-inflection,
prosodic isolation, and emotive force. The function is shared; the
category isn't.

A handful of interjections exceptionally take complements.
\textcite[1361fn]{HuddlestonPullum2002} analyse \mention{damn} in
\mention{damn these mosquitoes!} as the head of an interjection phrase
(IntjP) taking an NP complement~-- one of the very few cases where an
interjection projects beyond a single word. But even here the syntax is
constrained: no subject is present or intended, and there's no tense
inflection. This is better treated as a residue of verb syntax than as
full verbal behaviour~-- a point that matters for the recruitment pathways
discussed in Section~\ref{sec:stabilisers}. (The internal structure of lexicalized pairs like
\mention{oh boy} and \mention{holy cow} is less clear and won't be
resolved here.)

\textit{Inferential contribution.} Supplement function predicts prosodic
isolation and resistance to syntactic embedding:

\ea\label{ex:syntactic-predictions}
  \ea[*]{\mention{I think that damn, we're going to be late.}}
  \ex[*]{\mention{Damn and quickly, we left.}}
  \z
\z
It also predicts a semantic consequence: if the form is syntactically
non-integrated, its contribution is independent of the proposition
expressed by the host clause. Remove the interjection and the
truth conditions don't change~-- the hallmark of non-truth-conditional
meaning \citep[cf.][166]{potts2007}.

\subsection{Semantics}

English interjections characteristically don't refer. While nouns and verbs
typically pick out entities or events in the world, interjections express
internal states of their users \citep{leech2006}: anger (\mention{damn}),
disgust (\mention{eww}, \mention{yuck}), surprise (\mention{wow}), regret
(\mention{alas}), pain (\mention{ouch}), realization (\mention{eureka}).
Some interjections are more propositional than emotive~-- \mention{yes} and
\mention{no} affirm or deny propositions, and \mention{oh} can signal
recognition of new information rather than expressing feeling.
\textcite{wierzbicka1992} argues that interjections have genuine
semantic content, decomposable via Natural Semantic Metalanguage into
deictic primitives (\mention{I}, \mention{you}, \mention{now},
\mention{here}). This is not incompatible with non-referentiality as
used here: the claim is that interjections don't pick out entities in
the world, not that they're meaningless. Wierzbicka's explications
characterize the use-conditional content that
interjection\textsubscript{sem} (Section~\ref{sec:hpc-fieldrelative})
tracks. Non-referentiality is a cluster property, not a definitional
requirement.

The contrast with nouns is sharpest where both share a lexical source.
\mention{Jesus} as a noun refers to a person; \mention{Jesus!} as an
interjection doesn't. The interjection use has been
\enquote{bleached of [its] original meaning}~-- it no longer refers to
the entity the noun originally picked out \citep[72]{gehweiler2008}.
This bleaching is part of the recruitment pathway from noun to
interjection (Section~\ref{sec:stabilisers}).

\textit{Inferential contribution.} Non-referentiality predicts
distributional restrictions. A form that doesn't refer can't be made
definite (*\mention{the damn!}), can't be quantified over
(*\mention{every ouch}), and can't fill argument positions
(*\mention{Damn surprised me}, *\mention{There's damn}). These are
negative predictions, but they're non-trivial: they distinguish
interjections from nouns that have undergone only partial bleaching.

\subsection{Pragmatics}

Interjections typically carry exclamatory or imperative force: \mention{ouch!}
expresses pain, \mention{shh!} demands silence \citep[88]{quirk1985}. But
the pragmatic range extends beyond exclamation. Interjections greet
(\mention{hey}, \mention{hello}), mark agreement (\mention{yes},
\mention{amen}), indicate understanding (\mention{oh}, \mention{uh-huh}),
and signal backchanneling~-- the listener's ongoing attention to a speaker's
turn \citep{dingemanse2020, stivers2019}.

The most common interjections tagged in the Corpus of Global Web-Based
English \citep[GloWbE;][]{daviesfuchs2015} confirm this pragmatic breadth: \mention{yes}, \mention{no},
\mention{oh}, \mention{yeah}, \mention{hi}, \mention{hey}, \mention{wow},
\mention{hello}, \mention{ah}, \mention{ha}. The list is dominated by
interactional items~-- forms whose primary function is to manage the
conversation rather than to contribute propositional content.
\textcite{dingemanse2024} estimates that roughly one in seven
conversational turns is an interjection, and the most frequent are not
involuntary exclamations but interactional tools: continuers like
\mention{mm-hm} that scaffold extended turns, repair initiators like
\mention{huh?} that calibrate mutual understanding, and change-of-state
tokens like \mention{oh} that display evolving knowledge
\citep[cf.][]{norrick2009}.

The boundary with \term{routine formulae} is instructive.
\textcite[2--3]{coulmas1981} defines routine formulae as
\enquote{highly conventionalized prepatterned expressions whose occurrence
is tied to more or less standard communication situations}:
\mention{bye}, \mention{hello}, \mention{sorry}, \mention{thank you}.
\textcite[109--110]{ameka1992} argues these differ from interjections in
three ways: they have addressees, they're predictable responses to social
conventions, and they're always speech acts rather than mere reflections of
the speaker's mental state. \textcite[142]{wilkins1992} counters that these
differences make routine formulae \enquote{a distinct pragmatic and semantic
subtype of interjections}, not a separate category.

Routine formulae share some properties with interjections (prosodic
isolation, pragmatic function) but have additional characteristics
(social convention, addressee-directedness) that interjections don't
require. They sit at the boundary~-- a pattern that will turn out to be
theoretically significant (Section~\ref{sec:boundaries}).

\textit{Inferential contribution.} If a novel form carries emotive or
interpersonal force without contributing propositional content, you can
predict specific interactional behaviour: turn-initial or standalone
positioning, availability for backchanneling, and resistance to
embedding under reported speech (\mention{*She said that damn}).
These are predictions about distribution and sequential position, not
restatements of the meaning type. They follow because the mechanisms
that maintain emotive force (pragmatic conventionalization, social
indexing) also maintain the interactional niche.

\subsection{Co-occurrence, not necessity}\label{sec:cooccurrence}

These properties co-occur statistically, not necessarily. Every
characterization of interjections is hedged.
Descriptions across the literature characteristically hedge:
interjections \enquote{commonly} occupy their own intonation unit,
\enquote{typically} lack inflection, \enquote{usually} resist
syntactic integration, and \enquote{generally} express emotive meaning
\citep[cf.][]{ameka1992, wilkins1992}. No single property is necessary
and sufficient.

The pattern calls for explanation. The cluster holds together not because
a definition mandates it but because processes couple the properties:
prosodic isolation reinforces syntactic non-integration,
non-referentiality removes the need for argument structure, and
pragmatic conventionalization entrenches the emotive function.

The coupling is visible when it breaks. \mention{Damn} as an interjection
(\mention{Damn!}) has the full cluster: prosodic isolation, syntactic
non-integration, non-referentiality, emotive force. \mention{Damn} as an
attributive modifier (\mention{the damn car}) retains emotive force but
loses prosodic isolation and syntactic non-integration~-- it's embedded
inside an NP \citep[cf.][]{potts2007}. The properties don't erode
randomly; they erode together when the mechanisms that couple them
(here, prosodic-syntactic independence) are neutralized. A list of
co-occurring properties can't explain this pattern. A cluster whose
co-occurrence is maintained by identifiable processes can.

Table~\ref{tab:diagnostic-matrix} makes the cluster inspectable.
Each row is a construction (not a lexeme; see
Section~\ref{sec:intro}), and the columns show its profile against the
five cluster properties and three field-relative HPCs from
Section~\ref{sec:hpc-fieldrelative}. The table reveals where the
HPCs converge (standalone \mention{damn!}, \mention{oh}) and where
they decouple (attributive \mention{the damn car} is
interjection\textsubscript{sem} but not
interjection\textsubscript{syn}; \mention{mm-hm} is
interjection\textsubscript{int} but debatably
interjection\textsubscript{syn}). Boundary items (\mention{bye},
\mention{um}) share some but not all cluster properties~-- the pattern
that Section~\ref{sec:boundaries} explains as mechanism overlap.

\begin{table}[htbp]
\centering
\small
\caption{Diagnostic matrix: constructions against cluster properties and
field-relative HPCs. \checkmark\ = property present; ($\checkmark$) =
partial or debated; --- = absent. Sources: properties from
\textcite{ameka1992}, \textcite{quirk1985},
\textcite{HuddlestonPullum2002}; interjection\textsubscript{sem} from
\textcite{potts2007}; interjection\textsubscript{int} from
\textcite{stivers2019}, \textcite{dingemanse2020}.}
\label{tab:diagnostic-matrix}
\begin{tabular}{@{}l*{5}{c}c*{3}{c}@{}}
\toprule
 & \rotatebox{70}{Prosodic isol.}
 & \rotatebox{70}{Non-inflecting}
 & \rotatebox{70}{Synt.\ non-int.}
 & \rotatebox{70}{Non-referential}
 & \rotatebox{70}{Emotive/stance}
 & \phantom{x}
 & \rotatebox{70}{Intj\textsubscript{syn}}
 & \rotatebox{70}{Intj\textsubscript{sem}}
 & \rotatebox{70}{Intj\textsubscript{int}} \\
\midrule
\mention{gee!}
  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark
  && \checkmark & \checkmark & \checkmark \\
\mention{damn!}
  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark
  && \checkmark & \checkmark & \checkmark \\
\mention{damn these \ldots!}
  & \checkmark & ($\checkmark$) & ($\checkmark$) & \checkmark & \checkmark
  && ($\checkmark$) & \checkmark & --- \\
\mention{the damn car}
  & --- & --- & --- & --- & \checkmark
  && --- & \checkmark & --- \\
\mention{oh}
  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark
  && \checkmark & \checkmark & \checkmark \\
\mention{mm-hm}
  & \checkmark & \checkmark & \checkmark & \checkmark & ($\checkmark$)
  && ($\checkmark$) & --- & \checkmark \\
\mention{huh?}
  & \checkmark & \checkmark & \checkmark & \checkmark & ($\checkmark$)
  && ($\checkmark$) & --- & \checkmark \\
\mention{bruh}
  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark
  && \checkmark & \checkmark & \checkmark \\
\mention{bye}
  & \checkmark & \checkmark & \checkmark & ($\checkmark$) & ---
  && \checkmark & --- & ($\checkmark$) \\
\mention{um}
  & ($\checkmark$) & \checkmark & \checkmark & \checkmark & ---
  && ($\checkmark$) & --- & ($\checkmark$) \\
\bottomrule
\end{tabular}
\end{table}

The next two sections develop the theoretical framework
(Section~\ref{sec:hpc}) and the mechanisms themselves
(Section~\ref{sec:stabilisers}).

\section{HPC theory and why projectibility matters}\label{sec:hpc}

The property cluster documented above has a distinctive shape: statistically
reliable co-occurrence, no necessary and sufficient conditions, graded
boundaries, and hedged universals. Classical category theory can't
accommodate this. HPC theory can.

\subsection{The framework}

\textcite{boyd1999} proposed that natural kinds are not defined by
essences but sustained by homeostatic mechanisms. A kind is a property
cluster: a set of properties that co-occur reliably across instances.
The co-occurrence isn't accidental; it's maintained by mechanisms that
keep the properties clustered. And the clustering matters because it
supports induction~-- it lets you predict properties of new instances
from observed properties of old ones.

Following \textcite{boyd1999}, I adopt a three-part slogan: a
linguistic category is a profile of co-occurring properties, stabilized
by mechanisms, projectible relative to purposes. The first two legs
matter because they make the third possible. Projectibility is the
payoff: a category just is whatever supports reliable inference across
novel instances. The cluster is the visible trace; the inferential role
is the substance.

Consider prosodic isolation: if interjections lost it~-- if they were
integrated into the surrounding prosodic contour~-- would the category
hold together? A \term{stabiliser} is any process whose removal would
change the clustering~-- would cause the category to fragment, drift, or
dissolve \citep[cf.][]{boyd1999}. This yields the perturbation
test: if a proposed stabiliser can be neutralized and the cluster
remains intact, the stabiliser wasn't doing the work claimed for it.

Not everyone agrees that mechanisms are necessary. \textcite{slater2015}
argues that stable property clustering suffices for natural kindhood
without homeostatic mechanisms. I follow \textcite{boyd1999} in requiring identifiable stabilisers,
because they do explanatory work that bare clustering can't: they
generate the perturbation predictions examined in Section~\ref{sec:discussion}, and they
explain why the properties co-occur in the pattern they do rather than
some other.

The paper succeeds or fails on this question: does categorizing a novel
form as an interjection license non-trivial inferences for specific
purposes? If it merely lists co-occurring properties, it has described
a cluster, not demonstrated a kind.

\subsection{HPC vs.\ prototype theory}

Prototype theory \citep{rosch1975,taylor2003} was developed for exactly
the kind of data Section~\ref{sec:cluster} documents, and it handles it well. Graded
membership, better-and-worse exemplars, fuzzy boundaries via overlapping
similarity spaces~-- all of these are core predictions of prototype
theory, and all are confirmed by the interjection data. Prototype theory
also correctly predicts that boundary disputes (interjections vs.\
fillers, vs.\ routine formulae) should be persistent rather than
resolvable by definition. These are genuine explanatory achievements,
and HPC doesn't replace them.

HPC adds a different kind of question. Prototype theory describes the
fact that properties cluster but not why they cluster in the pattern
they do. It generates no
predictions about the order in which properties erode when items leave
the category (asymmetric fraying), whether apparent gradience
decomposes under social conditioning, or whether one property is more
central to cluster stability than others (hub-node asymmetry). These predictions, drawn from the framework's appendix of
falsifiable claims \citep[Appendix]{reynolds2025hpcbook}, are examined
for interjections in Section~\ref{sec:discussion}.

The perturbation test provides a further discriminator. The
interjection-as-HPC claim is falsifiable: if no proposed stabiliser's
removal would change the clustering, the \enquote{homeostatic} part is
empty. Prototype theory offers no
analogous test.

\subsection{Field-relative projectibility}\label{sec:hpc-fieldrelative}

\textcite[150]{boyd1999} argued that the naturalness of a kind is
relative to a disciplinary matrix: \enquote{A kind may be natural `from
the point of view of' some discipline or disciplinary matrix but not
`from the point of view of' another.} His example is jade, a natural
kind in gemology but not in geology (jadeite and nephrite are chemically
distinct). In linguistics, the parallel is the noun/name distinction: syntacticians track
\term{noun} (heads NPs, takes determiners, inflects for number);
semanticists track \term{name} (picks out individuals, resists
descriptive modification, creates referential opacity). Same words,
different HPCs, different mechanisms, different projectibility.

This matters for interjections. Multiple fields recognize a category
whose extension is roughly the same set of lexemes, but each field
tracks different properties, maintained by different mechanisms,
projectible for different purposes. I distinguish these with
subscripts:%
\footnote{The noun/name model \citep[Ch~7]{reynolds2025hpcbook} uses
  different words for the syntactic and semantic HPCs because both terms
  are established in their respective traditions. For interjections,
  every field uses the same word, so subscripts are the honest notation.
  An alternative would have been to adopt distinct terms from the
  relevant literatures~-- \term{expressive} from \textcite{potts2007}
  for the semantic HPC, \term{response token} from
  \textcite{stivers2019} or \term{liminal sign} from
  \textcite{dingemanse2020} for the interactional HPC.}

\begin{itemize}
\item \textbf{Interjection\textsubscript{syn}}~-- the syntactician's
  category (\textit{CGEL}, Quirk et al.). Distributional cluster: supplement
  function, non-inflecting, non-argument-taking. Maintained by
  prosodic-syntactic coupling. Projects syntactic behaviour.
\item \textbf{Interjection\textsubscript{sem}}~-- the semanticist's
  category, overlapping with \textcite{potts2007}'s \term{expressives}.
  Use-conditional meaning cluster: independent of at-issue content,
  nondisplaceable, perspective-dependent, descriptively ineffable
  \citep[166]{potts2007}. Maintained by semantic bleaching. Projects
  meaning type. Different extension: \mention{damn} in \mention{the damn
  car} is interjection\textsubscript{sem} but not
  interjection\textsubscript{syn}.
\item \textbf{Interjection\textsubscript{int}}~-- the interactional
  linguist's category, overlapping with \textcite{stivers2019}'s
  response tokens and \textcite{dingemanse2020}'s liminal signs.
  Sequential/stance cluster: turn-initial, backchannel, stance display.
  Maintained by conversational routines and social indexing. Projects
  interactional distribution. Different extension: \mention{mm-hm} is
  interjection\textsubscript{int} but debatably
  interjection\textsubscript{syn}.
\end{itemize}

The subscript notation makes explicit that these are distinct HPCs, not
subtypes of a single category. The traditional label
\enquote{interjection} names the intersection, and the paper's remaining
sections analyze that intersection from multiple angles.

\section{Stabilisers: recruitment and maintenance}\label{sec:stabilisers}

The property cluster documented in Section~\ref{sec:cluster} raises two
questions that co-occurrence alone can't answer: how do items enter the
cluster, and what keeps it coherent once they're in it? In what follows,
\enquote{the interjection cluster} refers to the intersection of the
three HPCs introduced in Section~\ref{sec:hpc-fieldrelative}~-- the
property profile shared by forms that belong to
interjection\textsubscript{syn}, interjection\textsubscript{sem}, and
interjection\textsubscript{int} simultaneously.
\textcite{boyd1999}'s framework requires both a diachronic account of
recruitment and a synchronic account of maintenance.

\subsection{Recruitment pathways}\label{sec:recruitment}

Section~\ref{sec:cluster} already pointed to three recruitment
pathways: semantic bleaching from nouns, syntactic deactivation from
verbs, and nonce coinage via onomatopoeia. The first two are analysed
in detail here; the third is taken up in
Section~\ref{sec:turnover}.

\subsubsection{Path 1: Religious noun to interjection}

\textcite{gehweiler2008} traces a four-stage trajectory from proper name to
primary interjection, using the development of \mention{gee} from
\mention{Jesus} as the central case:

\ea\label{ex:bleaching}
  \ea Stage 1: Proper name used as invocation (\mention{Jesus!} addressed
    to the deity)
  \ex Stage 2: Exclamatory use, no longer addressing the deity
    (\mention{Jesus!} as an expression of surprise)
  \ex Stage 3: Phonological modification (\mention{Jeez}, \mention{Gee})
  \ex Stage 4: Primary interjection (\mention{gee!}), fully detached from
    the source noun
  \z
\z
The stages don't install cluster properties one by one; prosodic
isolation is already present at stage~1, because a vocative like
\mention{Jesus!} occupies its own intonation unit (contrast the
integrated argument in \mention{Put your faith in Jesus}). What changes
is the content: non-referentiality arrives at stage~2, when the speaker
is no longer addressing the deity. Phonological divergence at stage~3
makes the etymological connection opaque. By stage~4, the form is a
standalone interjection that many speakers don't connect to its source.
The vocative is the gateway: it provides the prosodic isolation that
the bleaching pathway exploits. The prosodic slot was there first; the
semantic content drained out of it \citep[85]{gehweiler2008}.

This is a constrained trajectory, not a random walk. Each stage makes the
next more probable: once the invocation loses its referential force
(stage~2), the phonological connection to the source is the last link
to the nominal category, and weakening that link (stage~3) enables full
independence (stage~4). Gehweiler's corpus data confirms the
constraint: the phonological modifications (\mention{Jeez},
\mention{Gee}, \mention{Gosh}, \mention{Golly}) cluster in the
seventeenth and eighteenth centuries, with the transition from
invocation to primary interjection happening relatively rapidly once the
modifications appear \citep[80--85]{gehweiler2008}.

Two sub-mechanisms operate within this path.
\textcite[76]{gehweiler2008} distinguishes gradual Invited Inferencing
(speaker-based: the expressive use is conversationally implicated, then
conventionalized) from abrupt reanalysis (hearer-based: the hearer parses
the exclamation as a standalone unit rather than an invocation).
Both feed the same cluster, but they operate at different timescales and
involve different agents. That two distinct cognitive processes converge on
the same cluster profile is itself evidence for HPC: the cluster can be reached by different routes.

The constrained trajectory generates a specific prediction: partial
bleaching (stage~2) should be detectable by the absence of
downstream properties. A form at stage~2 should be non-referential but
should retain the source phonology and resist full prosodic independence.
This is testable and goes beyond historical description.

\subsubsection{Path 2: Verb to interjection}

A second trajectory enters the cluster via syntactic deactivation rather
than phonological modification. \mention{Damn} originates as a verb but
in its interjection use has lost subject and tense while retaining~-- in
a few constructions~-- an NP complement (\mention{damn these mosquitoes!}).
The complement retention shows that the bleaching is partial: the form
carries some verb properties into the interjection cluster. In \textit{CGEL}'s
terms, it's the head of an IntjP, not a VP
\citep[1361fn]{HuddlestonPullum2002}. The process is better described as
pragmaticalization with partial degrammaticalization: \mention{damn} as
a verb is a speech act (content meaning: condemning), while
\mention{damn!} as interjection\textsubscript{syn} is a stance marker
(procedural meaning: frustration). The loss of syntactic structure (subject, tense)
is degrammaticalization; the gain of expressive force is
pragmaticalization.

For other verb-derived interjections, the imperative is the gateway~--
just as the vocative is for Path~1. Imperatives already lack an overt
subject and can stand alone prosodically; the interjection use completes
the deactivation. \mention{Come on} and \mention{look} (in their
discourse-marker uses) have lost not just subject and tense but also
complement-taking ability. They function as standalone interjections
with no residual argument structure.
The contrast is instructive: \mention{damn} represents partial syntactic
deactivation (complement retained, subject and tense lost), while
\mention{come on} and \mention{look} represent full deactivation (no
complement, no argument structure at all). The degree of deactivation
correlates with the degree of integration into the interjection cluster.
This correlation generates a graded prediction: a form that retains
complement-taking ability should also retain other verb-like properties
(e.g., prosodic integration in some constructions), while a fully
deactivated form should show the complete interjection profile. The
mechanism doesn't just describe entry; it predicts the inferential
package available at each stage.

\subsubsection{Convergent pathways}

The two paths enter the cluster from different directions~-- one through
phonological modification of a bleached noun, the other through syntactic
deactivation of a verb~-- but both exploit a gateway construction that
already provides prosodic isolation: the vocative for Path~1, the
imperative for Path~2. And both converge on the same property profile.
The convergence is what HPC predicts: different mechanisms can recruit items
into the same kind, producing the same cluster of co-occurring properties
through different routes.

Both paths can be situated within \textcite{TraugottDasher2002}'s
subjectification trajectory (content~$>$~procedural,
nonsubjective~$>$~intersubjective), but they instantiate it differently.
Path~1 is grammaticalization in Traugott's technical sense: stage~1
(invocation) carries content meaning (addressing an entity) plus
subjective stance (reverence); stage~2 (exclamation) loses content and
gains procedural meaning (expressive force); stages~3--4 (phonological
modification) complete the shift to fully procedural, intersubjective
meaning~-- hearers interpret the form as a conventional expression, not
an invocation. Path~2, as described above, involves pragmaticalization with partial
degrammaticalization.

Onomatopoeia (Section~\ref{sec:turnover}) is categorically different: nonce coinages
enter the cluster with no prior lexical history~-- no bleaching, no
deactivation, no subjectification. Three diachronically distinct
processes (grammaticalization, pragmaticalization, nonce coinage)
converge on the same synchronic category. Grammaticalization theory
alone can't subsume all three. HPC theory can: what unifies the
pathways isn't the diachronic process but the cluster itself~--
convergent outcomes from different recruitment mechanisms, maintained by
the same synchronic stabilisers.

\subsection{Synchronic maintenance}\label{sec:maintenance}

Recruitment explains how items enter the cluster. But what keeps the
cluster coherent now? Three stabilisers operate synchronically.

\subsubsection{Prosodic isolation and syntactic non-integration}

These two properties reinforce each other, though the evidence is
stronger in one direction. Prosodic separation makes syntactic
integration harder: a form in its own intonation unit is less available
as a complement or modifier. \textcite[85]{gehweiler2008} provides direct
evidence for this direction: sentence-initial position of invocations
contributed to the emergence of the interjection~-- the prosodic slot
preceded and enabled the syntactic independence. The reverse direction
(syntactic independence licensing prosodic isolation) is plausible~-- a
form with no argument-structural requirements can stand alone
prosodically~-- but the historical evidence is primarily for the
prosody-to-syntax direction. The coupling is primarily prosody-driven,
with syntactic reinforcement.

\subsubsection{The non-referentiality ratchet}

Bleaching removes referential content, which removes the semantic
motivation for argument structure, which makes the form syntactically
free-standing. This is a one-directional cascade, not a feedback loop:
once argument structure is lost, there's no pressure to restore it,
because the form now occupies a different functional niche. It's a
ratchet rather than a cycle, but it contributes to cluster stability
just the same~-- it explains why items that enter the interjection
cluster don't drift back toward their source category.

\subsubsection{Social indexing (candidate stabiliser)}

Regional interjections survive because they do identity work.
\mention{Lah} in Singapore and Malaysia, \mention{yaar} in India and
Pakistan, \mention{haba} in Nigeria, and \mention{word} in African-American
English \citep{lauture2020} persist not because of their propositional
content (they have none) but because they mark in-group solidarity.
Social indexing is a candidate stabiliser, not yet a confirmed one. The
open question is whether it preserves the \emph{cluster} (maintaining
property co-occurrence) or merely preserves individual \emph{items}
(maintaining inventory). If the former, social indexing is a genuine HPC
stabiliser; if the latter, it's a lexical-persistence mechanism that
operates independently of the cluster.

A test: when a regional interjection is adopted by out-group speakers as
ironic quotation, does it retain its cluster properties or lose them? If
out-group \mention{lah} loses prosodic isolation or emotive force while
retaining its social-marking function, social indexing is maintaining the
inventory, not the cluster. If it retains the full profile, social
indexing is genuinely maintaining the HPC. The test is unexecuted.
The interjection cluster therefore has two solid synchronic
stabilisers (prosodic-syntactic coupling, the non-referentiality
ratchet) and one candidate (social indexing).

\subsection{Perturbation test}\label{sec:perturbation}

If the stabilisers are genuine, disrupting one should change the
clustering. The clearest cases involve interjections that gain verbal
morphology. \mention{Wow}, \mention{boo}, \mention{ooh}, and
\mention{shoo} are interjections with the full cluster profile:
prosodically isolated, non-inflecting, non-referential, syntactically
non-integrated. But all four have verb uses~-- \mention{she wowed the
audience}, \mention{the crowd booed}, \mention{they oohed and aahed},
\mention{I shooed them away} \citep[178--214]{biese1941}~-- and in those
uses the cluster comes apart asymmetrically across the three HPCs.
Interjection\textsubscript{syn} properties erode: the derived verb
inflects (\mention{wowed}, \mention{booing}), takes arguments
(\mention{wowed the audience}), and integrates syntactically.
Interjection\textsubscript{int} properties erode: the verb use isn't a
backchannel or standalone stance marker. But
interjection\textsubscript{sem} properties partially persist:
\mention{wow} as a verb still carries evaluative, speaker-oriented
meaning (\mention{she wowed them} expresses the speaker's assessment,
not a neutral description). The erosion isn't random; it tracks which
mechanisms are disrupted by the gain of morphology, and the three HPCs
come apart at different rates.

An interactional case shows a different erosion pattern. When
\mention{mm-hm}~-- an interjection\textsubscript{int} continuer~--
is replaced by a full clause (\mention{I agree}, \mention{I'm
listening}), the form gains propositional content and loses prosodic
independence: it integrates into the turn structure rather than
occupying its own intonation unit. Interjection\textsubscript{int}
properties erode (it's no longer a minimal backchannel), and
interjection\textsubscript{syn} erodes (it's now a clause, not a
supplement). But the interactional \emph{function} persists~-- the
speaker is still doing agreement or attention-marking. The erosion
pattern differs from the emotive cases above, because different
mechanisms are disrupted: gaining morphology disrupts prosodic-syntactic
coupling, while gaining propositional content disrupts the
non-referentiality ratchet.

The HPC-specific prediction here is not simply that syntactic properties
erode before semantic ones~-- any framework distinguishing syntax from
semantics could accommodate that. It's that the \emph{order} of erosion
tracks which mechanisms are disrupted: tight diagnostics (those
maintained by fewer mechanisms) erode before loose ones (those maintained
by multiple mechanisms). This ordering prediction is unavailable to
prototype theory, which has no mechanisms, and to construction grammar,
which has no perturbation test.

\subsection{Stability through turnover}\label{sec:turnover}

The interjection cluster is stable even as its membership changes:
individual members enter and leave, but the cluster persists.

The Corpus of Historical American English (COHA) illustrates this
turnover. Among words tagged as interjections, \mention{nay} ranks in
the top~10 in the 1820s but falls to negligible frequency by the 2010s;
\mention{yeah}, entirely absent from the 1820s data, is among the five
most frequent interjection-tagged forms by the 2010s.%
\footnote{Based on part-of-speech--tagged frequencies in COHA \citep{davies2010}.
  Frequencies are normalized per million words. The interjection tag
  captures a conventional inventory, not a theoretically motivated one;
  the point here is the turnover pattern, not the tagging decisions.}
The items change; the category doesn't.

Regional variation confirms the same point from a geographic angle. The
GloWbE corpus \citep[1.9~billion words across 20
English-speaking countries]{daviesfuchs2015} shows interjections distributed unevenly
across varieties~-- \mention{lah} almost exclusively in Singapore and
Malaysia, \mention{yaar} in India and Pakistan, \mention{haba} in
Nigeria~-- but the cluster profile (prosodic isolation,
non-referentiality, emotive force, syntactic non-integration) is
recognizable across all of them.

The cluster is also continuously replenished via onomatopoeia. Nonce
interjections can be freely coined \citep[74]{quirk1985}, and new coinages
immediately adopt the cluster profile: they're prosodically isolated,
non-referential, emotively laden, and syntactically non-integrated.

This generates a testable prediction: a newly conventionalized
interjection should exhibit the full cluster profile without having
traversed any of the recruitment pathways. Recent forms like
\mention{bruh}, \mention{yeet}, and \mention{oof} bear this out~-- all
are prosodically isolated, non-inflecting, non-referential, and
syntactically non-integrated, despite entering the language through
different routes (phonological reduction, coinage, onomatopoeia
respectively). The cluster persists even as its members turn over.

\textcite{dingemanse2020} characterizes such items as \term{liminal signs}~--
signs at the boundary between sound and speech, whose liminality is
functional, not defective. Cultural evolutionary processes \enquote{sample
the landscape of possible resources}. HPC provides the framework for
that landscape: a property space with convergent maintenance by the
stabilisers above.

\section{Projectibility: what the category lets you predict}\label{sec:projectibility}

Section~\ref{sec:cluster} showed what each property, individually, lets
you infer. Section~\ref{sec:stabilisers} identified the mechanisms that
couple those properties. This section asks a different question: what
does the package license, and for whom? From here on, bare
\enquote{interjection} refers to the intersection of the three HPCs
unless a subscript specifies otherwise. The paper's criterion of success
(Section~\ref{sec:hpc}) was that categorizing a novel form as a member
of this intersection should license non-trivial inferences for specific
purposes.

\subsection{Partial-observation inference}

A circular account of projectibility says: if X is an interjection, it
has interjection properties. That's trivially true of any category. HPC
projectibility is different. It says: observing some cluster
properties in a novel form licenses inference to other properties you
haven't yet observed.

Consider \mention{bruh}, a recently conventionalized form. A speaker
encountering it for the first time observes that it's prosodically
isolated and carries stance meaning. From those two observations alone,
the three interjection HPCs license a battery of further predictions:

\ea\label{ex:bruh}
  \ea Interjection\textsubscript{syn}: it won't inflect
    (*\mention{bruhs}, *\mention{bruhed}), will resist syntactic
    integration (*\mention{I think that bruh, we're late}), and won't
    take a determiner or function as an argument (*\mention{the bruh},
    *\mention{Bruh surprised me}).
  \ex Interjection\textsubscript{int}: it will be available for
    turn-initial or standalone use.
  \ex Interjection\textsubscript{sem}: its expressive meaning will be
    use-conditional and speaker-oriented.
  \ex Diachronic: its etymological source is a content word
    (\mention{brother}) that has undergone phonological reduction and
    semantic bleaching.
  \z
\z
These predictions vary in strength, and the asymmetry matters. The
interjection\textsubscript{syn} predictions (a) are close to
entailments given English morphology: non-inflection and
non-referentiality straightforwardly block determiner and argument
functions. These aren't inferences \emph{from} HPC; they're
consequences of the properties themselves. The
interjection\textsubscript{int} prediction (b) involves a genuine
cross-domain leap from phonology to interactional distribution. The
diachronic prediction (d) is the strongest: it predicts etymological
trajectory from synchronic properties. But (d) is available only to
linguists who recognize the category, not to naive speakers~-- it is
projectibility for the discipline, not for the language user. That
asymmetry is legitimate under \textcite[150]{boyd1999}'s
field-relativity, but it should be visible, not hidden.
All four succeed because the maintaining mechanisms
couple the cluster: the non-referentiality ratchet (Section~\ref{sec:maintenance})
connects loss of referential content to loss of argument structure to
syntactic independence; the prosodic-syntactic coupling connects
syntactic independence to prosodic isolation. The inferences are grounded
in mechanisms, not in definitions.

\subsection{Field-relative projectibility}\label{sec:proj-fieldrelative}

Different fields track categories with roughly the same extension but
different mechanisms and different inferential payoffs. Following
\textcite[150]{boyd1999} on the discipline-relativity of natural kinds,
the three interjection HPCs introduced in Section~\ref{sec:hpc-fieldrelative} pay off here.

\textbf{Interjection\textsubscript{syn}} projects syntactic behaviour.
If a novel form is categorized as interjection\textsubscript{syn}, a
syntactician can predict: supplement function, resistance to
\mention{that}-complementation, resistance to coordination with
non-interjections, no argument structure (or at most a bare NP
complement, as with \mention{damn}). These are distributional
predictions grounded in the prosodic-syntactic coupling mechanism from
Section~\ref{sec:maintenance}.

\textbf{Interjection\textsubscript{sem}} projects meaning type. If a
form is categorized as interjection\textsubscript{sem}~-- overlapping
with what \textcite{potts2007} calls \term{expressives}~-- a semanticist
can predict: use-conditional rather than truth-conditional contribution,
speaker-oriented attitude (perspective dependence), nondisplaceability
(the attitude is predicated of the utterance situation), and descriptive
ineffability \citep[166]{potts2007}. Potts defines expressives as a
semantic type in a multi-dimensional logic; I treat these four
properties as a cluster sustained by the bleaching mechanism from
Section~\ref{sec:recruitment}, not as a stipulative definition. The HPC reading is
supported by graded membership: \mention{damn} in \mention{the damn car}
has perspective dependence and use-conditionality but is arguably
displaceable in some reported-speech contexts~-- it has some but not all
of the cluster properties. This is the HPC signature, not a problem
for the analysis. The extension differs from
interjection\textsubscript{syn}: attributive \mention{damn} is
interjection\textsubscript{sem} but not
interjection\textsubscript{syn}~-- \textcite{potts2007} calls it an
expressive attributive adjective. Standalone \mention{damn!} is both.
The same lexeme belongs to different categories depending on
construction, which is exactly the pattern the subscript architecture
is designed to capture. Independent support comes from
\textcite{gutzmann2019}, who argues that expressivity is a syntactic
feature with agreement and movement reflexes~-- not merely a pragmatic
overlay. His case studies of expressive adjectives, intensifiers, and
vocatives show that the properties tracked by
interjection\textsubscript{sem} have grammatical structure, reinforcing
the claim that interjection\textsubscript{sem} is a genuine HPC, not
a pragmatic residue.

\textbf{Interjection\textsubscript{int}} projects interactional
distribution. If a form is categorized as
interjection\textsubscript{int}~-- overlapping with
\textcite{stivers2019}'s response tokens and
\textcite{dingemanse2020}'s liminal signs~-- a conversation analyst can
predict: turn-initial or standalone positioning, own intonation contour,
stance display rather than propositional content, and availability for
backchanneling. These are interactional predictions grounded in
conversational routines and social indexing. The extension differs from
interjection\textsubscript{syn}: \mention{mm-hm} is
interjection\textsubscript{int} (a backchannel) but its status as
interjection\textsubscript{syn} is debated
\citep{oconnell2005}.

These are not three descriptions of the same category. They are three
HPCs with overlapping but non-identical extensions, maintained by
different mechanisms, projectible for different purposes. The overlap is
explained by the fact that some mechanisms (prosodic isolation,
non-referentiality) contribute to multiple HPCs simultaneously. The
non-overlap is explained by the fact that other mechanisms are
field-specific: structural analogy (the extension of existing
syntactic patterns to new members by formal similarity) maintains
interjection\textsubscript{syn} but not
interjection\textsubscript{int}; conversational routines maintain
interjection\textsubscript{int} but not
interjection\textsubscript{syn}. A syntactician and a conversation
analyst can carve the same words differently without either being wrong.

The architecture generates a cross-HPC prediction: because some
mechanisms contribute to all three HPCs simultaneously, membership in
one should predict membership in the others. A novel form categorized as
interjection\textsubscript{syn} (prosodically isolated, non-inflecting,
syntactically non-integrated) should also tend to be
interjection\textsubscript{sem} (use-conditional, speaker-oriented) and
interjection\textsubscript{int} (turn-initial, available for
backchanneling). This is the default, and it holds for most interjections
(\mention{wow}, \mention{ouch}, \mention{hey}). The interesting cases
are where the prediction fails: attributive \mention{damn} is
interjection\textsubscript{sem} but not
interjection\textsubscript{syn}; \mention{mm-hm} is
interjection\textsubscript{int} but debatably
interjection\textsubscript{syn}. Each failure identifies a point where
the mechanisms decouple~-- and the pattern of decoupling is itself
predicted by which mechanisms are field-specific.

The decomposition also isn't just notational. It generates predictions
that the individual frameworks can't produce in isolation. The three
constructions of \mention{damn} illustrate this. For standalone
\mention{damn!}, all three HPCs agree: the form is prosodically
isolated, syntactically non-integrated, use-conditional, and available
as a stance marker. For \mention{damn these mosquitoes!}, the picture
is mixed: interjection\textsubscript{syn} holds (it heads an IntjP)
but with residual verb-like complement-taking;
interjection\textsubscript{sem} holds (the meaning is expressive);
interjection\textsubscript{int} is weakened (it's not a backchannel or
turn-management device). For attributive \mention{the damn car}, they
diverge further: interjection\textsubscript{syn} fails entirely (the
form is syntactically integrated as an adjective), but
interjection\textsubscript{sem} holds (use-conditional, perspective-
dependent). Three constructions, three different patterns of
convergence and divergence. Neither \textit{CGEL}'s syntax nor Potts's
semantics alone predicts this gradient. The subscript architecture
does, because it tracks which mechanisms are active in each
construction.

\subsection{Category vs.\ function}

A question from the \textit{CGEL} tradition sharpens the projectibility point:
what does the category label
\enquote{interjection\textsubscript{syn}} add beyond the function label
\enquote{supplement}? Supplements of any category~-- NPs, AdjPs,
relative clauses~-- share prosodic isolation and syntactic
non-integration. But interjection\textsubscript{syn} adds a package of
predictions that no other category-in-supplement-function supports:

\begin{itemize}
\item \textbf{Non-referentiality.} Supplements can be referential
  (\mention{a university professor, Dr.~Brown, \ldots}); interjections
  characteristically can't.
\item \textbf{Emotive/interpersonal force.} Not all supplements express
  stance; interjections characteristically do.
\item \textbf{Free coinage.} No other category in supplement function
  can be freely coined via onomatopoeia.
\item \textbf{Non-inflection.} NP supplements inflect freely;
  interjections don't.
\item \textbf{Diachronic trajectory.} Knowing something is a supplement
  makes no prediction about its etymology; knowing it's
  interjection\textsubscript{syn} does (bleaching paths, phonological
  reduction, constrained trajectories).
\end{itemize}

The function label tells you about syntactic integration. The category
label tells you about inference~-- it licenses a package of predictions
that the function label alone can't provide. This distinction matters
beyond the \textit{CGEL} tradition: any framework that distinguishes what kind of
thing an expression is from what job it's doing in a particular clause
will face the same question, and the HPC answer is the same. The
category earns its keep by licensing inferences that the function
doesn't.

\subsection{Interjections as a pragmatic kind}

The subscript architecture reveals something about HPC theory itself.
Not all kinds project in the same domain. Nouns project
morphosyntactically: they take determiners, inflect for number, fill
argument slots. Categorizing something as a noun tells you almost
nothing about its pragmatics~-- nouns aren't a pragmatic category.
For interjections, the situation is reversed. The three HPCs overlap,
and the richest projectibility belongs to
interjection\textsubscript{int}~-- the interactional cluster.
\textcite{dingemanse2024} reaches the same conclusion from interactional
data: the most frequent and functionally central interjections are not
emotive exclamations but interactional tools~-- continuers, repair
initiators, change-of-state tokens~-- whose primary projectibility is
about sequential position and conversational management. The forms at
the intersection of all three HPCs are what I call a \term{pragmatic
kind}: their strongest inferences are about how they will be used, not
what they will denote.

Why pragmatic? Because mechanisms generate predictions within the domains
they operate in. The prosodic-syntactic coupling from Section~\ref{sec:maintenance}
predicts prosodic and syntactic behaviour~-- not morphological paradigms
or truth-conditional content. Social indexing predicts persistence in
identity-marking communities~-- not inflectional regularity. The
mechanisms that maintain the three interjection HPCs are prosodic,
interactional, and social, so the predictions they underwrite are in
those domains. This isn't a stipulation; it's a consequence of which
mechanisms happen to converge. Noun projectibility is morphosyntactic
because the maintaining mechanisms (agreement, structural analogy,
discourse frequency) operate in morphosyntactic domains.

This connects to the content/procedural distinction. Interjections are
procedural, not conceptual \citep{TraugottDasher2002}: their
projectibility is about use, not denotation. The
interjection\textsubscript{sem} (expressive) cluster is the semantic
reflex: use-conditional meaning is the semantic signature of procedural
content. The parallel with other domain-specific kinds confirms the
pattern. \term{Animacy} projects grammatically (argument realization,
reference tracking); \term{colour} projects ecologically (edibility,
toxicity) but not grammatically~-- no language has colour-based verb
agreement \citep[cf.][]{seifart2010}. In each case, the domain of
projectibility follows from the maintaining mechanisms.

A functionalist might object: if the richest projectibility is pragmatic,
why keep interjection\textsubscript{syn} in the grammar at all?
The answer is that interjection\textsubscript{syn}, thin as it is, is a
genuine HPC with its own stabilisers and its own perturbation test
(Section~\ref{sec:perturbation}). Neutralize prosodic isolation and the form is
reclassified. The syntactic category does real distributional work~--
it predicts supplement function, embedding resistance, and coordination
restrictions~-- even if those predictions are less rich than the
pragmatic ones. A thin syntactic HPC is still a syntactic HPC.
Boyd's framework doesn't require all kinds to be equally thick; it
requires that they support reliable inference, and
interjection\textsubscript{syn} does.

\section{Discussion}\label{sec:discussion}

\subsection{Empirical predictions: HPC vs.\ prototype}

Section~\ref{sec:projectibility} asked what categorizing a novel form as an interjection lets
you predict about that form. This section shifts from instance-level
projectibility to category-level predictions: what does HPC predict about
the interjection intersection as a whole that prototype theory doesn't?

Prototype theory accommodates the interjection data~-- graded
membership, fuzzy boundaries, better-and-worse exemplars~-- and has been
applied to lexical categories since \textcite{rosch1975}. HPC doesn't
replace prototype theory's descriptive achievements. It adds something
prototype theory can't provide: mechanism-derived predictions with
named defeat conditions. Three are testable for interjections.

\textbf{Asymmetric fraying.} If items drift away from the
interjection cluster, tight diagnostics (non-inflection, complement
resistance) should erode before loose ones (prosodic isolation).
\textit{Defeat:} diagnostics erode in no principled order. The
evidence already in hand is consistent with the prediction. \mention{Damn} retains
complement-taking ability (\mention{damn these mosquitoes!}) but has
lost subject and tense~-- the tighter verbal diagnostics eroded first,
while the looser one (complement frame) persisted. \mention{Come on}
and \mention{look} have lost everything, including complements,
consistent with a trajectory where tight diagnostics go first and loose
ones follow. The perturbation cases from
Section~\ref{sec:perturbation} point in the same direction:
\mention{wow}~$>$~\mention{wowed} gains morphology and loses
interjection\textsubscript{syn} before losing
interjection\textsubscript{sem}. Three cases are preliminary, not
statistical, but the ordering is consistent and the prediction is
falsifiable. Prototype theory predicts no particular order of erosion.

\textbf{Conditioning recovers structure.} Apparent gradience at the
interjection boundary (fillers, routine formulae) should decompose when
conditioned on register, dialect, or community. \textit{Defeat:}
conditioned and pooled models perform identically. A preliminary test
uses GloWbE frequency data \citep{daviesfuchs2015}: for the 10 most
frequent interjection-tagged items plus regional and boundary items,
frequency-per-million was extracted by country (20 countries) and the
coefficient of variation (CV = SD/mean) was computed across countries.

The results show three tiers of geographic conditioning. Regional
interjections (\mention{lah}, \mention{yaar}, \mention{haba}) are
near-categorical: \mention{lah} occurs at 9.34 per million in Malaysia
and 6.72 in Singapore but near zero elsewhere; \mention{yaar} at 2.88
in Pakistan and 2.33 in India but near zero elsewhere; \mention{haba}
at 8.77 in Nigeria but near zero elsewhere. Core interjections show
intermediate but structured variation: \mention{ah} has the highest
CV (0.66), with Singapore (60.45) and Jamaica (53.20) at eight times
the rate of Kenya (7.62). Even broadly distributed items like
\mention{yeah} (CV 0.49) show a 6:1 ratio between the United States
(103.78) and Tanzania (16.87). Only \mention{yes} and \mention{no}
approach uniformity (CVs of 0.20--0.22). Boundary items confirm the
pattern: the fillers \mention{um} (CV 0.47) and \mention{uh} (CV 0.55)
and the routine formula \mention{bye} (CV 0.47) all show structured
geographic variation.

Conditioning on country recovers structure at every level of the
interjection landscape. The apparent gradience between core
interjections, boundary items, and non-interjections is not noise; it
is structured by variety. A Bayesian multilevel model (negative binomial with crossed random
effects for item and country) confirms that adding country as a
grouping factor substantially improves prediction over an item-only
baseline: the country-level random-effect standard deviation is 0.44
(95\% CrI: 0.28--0.65), and the conditioned model outperforms the
unconditional model by 22 ELPD units on leave-one-out
cross-validation. Full model specification and results are in the
supplementary materials.\footnote{Available at
  \url{https://github.com/BrettRey/interjections-hpc}.}
Two methodological caveats: GloWbE's
part-of-speech tagger, trained on inner-circle English, fails to tag
regional interjections (\mention{yaar} and \mention{haba} receive no
interjection tag at all), which means the tagged frequencies
underestimate regional items; and the \mention{ha} data are noisy
because the tagger captures laughter representations
(\mention{HA! HA!}) alongside genuine interjection uses.

\textbf{Hub-node asymmetry.} One property should be more central to
cluster stability than others. Candidates: prosodic isolation or
non-referentiality. Removing the hub should destabilize the cluster more
than removing a peripheral property. \textit{Defeat:} all properties
contribute equally. The perturbation test in Section~\ref{sec:perturbation} provides
preliminary evidence. When prosodic isolation is neutralized~-- as when
\mention{pop} in \mention{pop went the cork} is syntactically
integrated and reclassified as an adverb
\citep[152]{meinard2015}~-- reclassification is immediate. Restoring
partial referentiality, by contrast, allows the form to hover between
noun and interjection uses (as with \mention{God}). This suggests
prosodic isolation may be the hub, but the evidence is indirect.

Each prediction has a named defeat condition. Prototype theory generates
none of them~-- it describes the gradience but doesn't predict the
pattern of erosion, the trajectory of entry, or the structure hidden in
apparent variation.

\subsection{Interjections as a thin-but-genuine HPC}

Where do interjections sit on the word-class landscape? Not in the
wastebasket.
\term{Adverb} is widely regarded as a heterogeneous residual
category~-- a label with no shared mechanism and no convergence. Knowing
something is an adverb tells you it isn't a noun, a verb, or an
adjective, and almost nothing else.

Interjections are different. They aren't a robust skeleton category like
nouns or verbs (full mechanism convergence, deep entrenchment, early
acquisition). But they are a genuine thin HPC, comparable to
adjectives \citep[Ch~11]{reynolds2025hpcbook}. The parallel is worth
sketching. Adjectives, like interjections, have fewer stabilisers than
nouns or verbs, graded boundaries (with adverbs, participles, nouns),
and cross-linguistic instability as an independent category~-- many
languages lack a distinct adjective class. When English
adjectives gain nominal morphology (e.g., nominalization:
\mention{the poor}, \mention{the impossible}), they lose gradability
and predicative use before losing attributive use~-- the same
tight-before-loose erosion pattern that interjections show. Both
categories resist classical definition while supporting reliable
inference; both are maintained by a small set of mechanisms rather than
by the dense convergence that sustains nouns and verbs. The
category is maintained, not merely labelled.
Interjection\textsubscript{syn} is thin on its own; the richness comes
from the intersection of three overlapping HPCs
(interjection\textsubscript{syn}, interjection\textsubscript{sem},
interjection\textsubscript{int}), as Section~\ref{sec:projectibility} showed.

\subsection{Boundary disputes as mechanism overlap}\label{sec:boundaries}

The graded boundaries that embarrassed classical category theory~--
interjections vs.\ nouns, verbs, adverbs, fillers, routine formulae~--
are predictions of HPC, not anomalies to be explained away. Each
boundary reflects partial mechanism overlap.

Routine formulae (\mention{bye}, \mention{hello}, \mention{sorry}) share
prosodic isolation and pragmatic function with interjections but add
addressee-directedness and social predictability
\citep[109--110]{ameka1992}. They're maintained by additional mechanisms
(social convention, politeness norms) that prototypical interjections
don't require. In HPC terms, they share enough stabilisers to cluster
nearby but not enough to sit at the centre.

Fillers (\mention{um}, \mention{uh}) share prosodic distinctiveness and
syntactic non-integration but differ in emphasis, pause distribution,
and citation-introduction ability \citep{oconnell2005}. They occupy a
neighbouring region of the property space, maintained by partially
overlapping mechanisms (processing-pressure management rather than
emotive expression).

The Ameka-Wilkins debate is resolved: routine formulae are partial
members of the interjection cluster, not a separate category
\citep[142]{wilkins1992} and not fully inside it either
\citep[109]{ameka1992}. Both analysts were right about the data. The
disagreement was about the framework, and HPC provides the framework
that accommodates both observations.

\section{Conclusion}\label{sec:conclusion}

English interjections are a thin-but-genuine HPC: a property cluster
maintained by both diachronic recruitment pathways
(grammaticalization, pragmaticalization, nonce coinage) and synchronic
stabilisers (prosodic-syntactic coupling, the non-referentiality
ratchet, social indexing).

The category that Jespersen dismissed and Huddleston \& Pullum set aside
turns out to be a strong test case for HPC theory. Three diachronically
distinct processes converge on the same synchronic cluster. The
perturbation test suggests that the stabilisers are genuine: removing
one produces predictable, asymmetric category shifts.

What does categorizing something as an interjection let you predict?
Primarily pragmatic inferences~-- about prosodic behaviour,
interactional distribution, and stance expression~-- grounded in
identifiable mechanisms and reliable across novel instances. The
field-relative decomposition (interjection\textsubscript{syn},
interjection\textsubscript{sem}, interjection\textsubscript{int}) shows
that the same lexemes participate in distinct HPCs for different fields.
The traditional label names the intersection~-- a pragmatic kind, thin
syntactically but rich interactionally.

For historical pragmatics specifically, the payoff is this: interjection
formation involves at least three distinct diachronic processes
(grammaticalization via semantic bleaching, pragmaticalization via
syntactic deactivation, and nonce coinage via onomatopoeia), and HPC
theory explains why they converge on a recognizable category despite
their different trajectories. \textcite{gehweiler2008}'s analysis of
\mention{gee!} is a single case study; the present paper extends it to
a general claim about how the interjection intersection is recruited and
maintained. The mechanisms are diachronic; the stability they produce is
the kind of thing historical pragmatics is for.

Of the three predictions examined in Section~\ref{sec:discussion},
conditioning recovers structure is now supported by GloWbE data:
geographic conditioning decomposes apparent gradience across the
interjection landscape. Hub-node asymmetry remains untested: prosodic
isolation should destabilize the cluster more than other properties when
removed, and the perturbation evidence is suggestive but indirect. If
the hub-node prediction fails, the HPC claim is weakened but not
refuted. If it holds, interjections join the growing list of linguistic
categories whose structure is better explained by mechanisms than by
definitions.

\section*{Acknowledgements}

I compiled empirical material for the Wikipedia article
\enquote{English interjections}, of which I am the primary author. This
paper reanalyses that material through an HPC lens.

Supplementary materials (data and analysis code for the GloWbE
conditioning test) are available at
\url{https://github.com/BrettRey/interjections-hpc}.

I drafted this paper with the assistance of Claude (Anthropic). I have
reviewed and revised all content and take full responsibility for the
final text.

\newpage
\printbibliography

\end{document}
